```{r}
# 1. Sleep Hours vs Stroop Task Reaction Time
plot(sleep_df$Sleep_Hours, sleep_df$Stroop_Task_Reaction_Time,
     main = "Sleep Hours vs Stroop Reaction Time",
     xlab = "Sleep Hours",
     ylab = "Stroop Reaction Time (sec)",
     col = "blue", pch = 19)



```
```{r}
# 2. Sleep Hours vs N-Back Accuracy
plot(sleep_df$Sleep_Hours, sleep_df$N_Back_Accuracy,
     main = "Sleep Hours vs N-Back Accuracy",
     xlab = "Sleep Hours",
     ylab = "N-Back Accuracy (%)",
     col = "darkgreen", pch = 19)

```
```{r}
# 3. Sleep Quality Score vs Daytime Sleepiness
plot(sleep_df$Sleep_Quality_Score, sleep_df$Daytime_Sleepiness,
     main = "Sleep Quality vs Daytime Sleepiness",
     xlab = "Sleep Quality Score",
     ylab = "Daytime Sleepiness Score",
     col = "purple", pch = 19)

```
```{r}
# 4. Sleep Hours vs Stress Level
plot(sleep_df$Sleep_Hours, sleep_df$Stress_Level,
     main = "Sleep Hours vs Stress Level",
     xlab = "Sleep Hours",
     ylab = "Stress Level",
     col = "red", pch = 19)
```

Here we are able to quickly visualize what numeric variables relate to each other and the strength of these relationships. As the correlation plot shows the variables with positively correlated relationships are those in blue, the darker the blue color of the circles the stronger that positive relationship is. And for the red colored circles is a negative correlated for example in our data set it is apparent that the less sleep you get the less the N_Back_Accuracy score is(memory capabilities/attention). 
```{r}

#Visualize correlation between the numeric variables in our data set

install.packages("corrplot") 
library(corrplot)
numeric_var <- c( "Sleep_Hours", "Stroop_Task_Reaction_Time", "N_Back_Accuracy", "PVT_Reaction_Time" , "BMI", "Sleep_Quality_Score", "Stress_Level" )
cor_mat <- cor(sleep_df[, numeric_var])
corrplot(cor_mat)




```
